Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-11-28 formally.tex
ran the tex files through aspell and found a few minor typos

  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
    then there exists some $\bar\alpha$ such that
    $\bar\alpha~\mathsf{movements}$ nand
    $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.

>>>>

  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
    then there exists some $\bar\alpha$ such that
    $\bar\alpha~\mathsf{movements}$ and
    $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.

### 2016-11-28 hazelnut-popl17.tex
fixed awkward gaps
ran the tex files through aspell and found a few minor typos

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 \begin{tabular}{ccc}
 	\begin{tabular}{c}
 $^1$Carnegie Mellon University (USA)\\
 \textsf{\hphantom{$^1$}\{comar, iev, aldrich\}@cs.cmu.edu}
 \end{tabular} & 
 \begin{tabular}{c}
 $^2$Oregon State University (USA)\\
 \textsf{\hphantom{$^1$}hiltonm@eecs.oregonstate.edu}
 \end{tabular} & 
  \begin{tabular}{c}
 $^3$University of Colorado Boulder (USA)\\
 \textsf{\hphantom{$^1$}matthew.hammer@colorado.edu}
 \end{tabular}
 \end{tabular}
}
{
 % \vspace{-12px}
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

@@

An intermediate approach would be to 
layer unification-based type generation features atop the bidirectional system. This would amount to  interpretting type holes as unification
variables. For a simple calculus,
e.g. the STLC upon which Hazelnut is based, type inference for complete expressions is known to be
decidable, so type holes could be instantiated automatically once the expression that they appear within has been constructed. It would also be possible to flag expressions for which there does not exist any way to fill the type holes. In more complex settings, e.g. in a 
dependently typed language, a partial decision procedure may still be
useful in this regard, both at edit-time and (just prior to)
run-time. Indeed, text editor modes for dependently typed proof assistants, e.g. for Agda,
attempt to do exactly this for indicated ``type holes'' (and do not always
succeed.)

@@

One interesting avenue of future work is to elaborate expression holes to
CMTT's closures, i.e. CMTT terms of the form
$\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable
associated with each hole and $\text{id}(\Gamma)$ is the explicit identity
substitution. This would allow us to evaluate expressions with holes such
that the closure ``accumulates'' substitutions explicitly. When evaluation
gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped
with a notion of progress under a non-empty modal context), it would then
be possible for the programmer to choose holes from amongst the visible
holes (which may have been duplicated) to edit in their original
context. Once finished, the CMTT hole instantiation operation, together
with a metatheorem that establishes that reduction commutes with
instantiation, would enable an ``edit and resume'' feature with a clear
formal basis. This notion of reduction commuting with instantiation has
also been studied in other
calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running
program also has connections to less formal work on ``live programming''
interfaces \cite{burckhardt2013s,lamdu}.

@@

\subsubsection{Collaborative Programming}
Finally, we did not consider any aspects of \emph{collaborative
programming}, such as a packaging system, a differencing algorithm for use
in a source control system, support for multiple simultaneous focii for
different users, and so on. These are all interesting avenues for future
work.

>>>>

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 \begin{tabular}{ccc}
 	\begin{tabular}{c}
 $^1$Carnegie Mellon University, USA\\
 \textsf{\hphantom{$^1$}\{comar, iev, aldrich\}@cs.cmu.edu}
 \end{tabular} & 
 \begin{tabular}{c}
 $^2$Oregon State University, USA\\
 \textsf{\hphantom{$^1$}hiltonm@eecs.oregonstate.edu}
 \end{tabular} & 
  \begin{tabular}{c}
 $^3$University of Colorado Boulder, USA\\
 \textsf{\hphantom{$^1$}matthew.hammer@colorado.edu}
 \end{tabular}
 \end{tabular}
}
{
 % \vspace{-12px}
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

@@

An intermediate approach would be to 
layer unification-based type generation features atop the bidirectional system. This would amount to  interpreting type holes as unification
variables. For a simple calculus,
e.g. the STLC upon which Hazelnut is based, type inference for complete expressions is known to be
decidable, so type holes could be instantiated automatically once the expression that they appear within has been constructed. It would also be possible to flag expressions for which there does not exist any way to fill the type holes. In more complex settings, e.g. in a 
dependently typed language, a partial decision procedure may still be
useful in this regard, both at edit-time and (just prior to)
run-time. Indeed, text editor modes for dependently typed proof assistants, e.g. for Agda,
attempt to do exactly this for indicated ``type holes'' (and do not always
succeed.)

@@

One interesting avenue of future work is to elaborate expression holes to
CMTT's closures, i.e. CMTT terms of the form
$\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable
associated with each hole and $\text{id}(\Gamma)$ is the explicit identity
substitution. This would allow us to evaluate expressions with holes such
that the closure ``accumulates'' substitutions explicitly. When evaluation
gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped
with a notion of progress under a non-empty modal context), it would then
be possible for the programmer to choose a hole from the visible
holes (which may have been duplicated) to edit in their original
context. Once finished, the CMTT hole instantiation operation, together
with a metatheorem that establishes that reduction commutes with
instantiation, would enable an ``edit and resume'' feature with a clear
formal basis. This notion of reduction commuting with instantiation has
also been studied in other
calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running
program also has connections to less formal work on ``live programming''
interfaces \cite{burckhardt2013s,lamdu}.

@@

\subsubsection{Collaborative Programming}
Finally, we did not consider any aspects of \emph{collaborative
programming}, such as a packaging system, a differencing algorithm for use
in a source control system, support for multiple simultaneous cursors for
different users, and so on. These are all interesting avenues for future
work.
